---
title: "The Anatomy of a Hit - What Drives Song Popularity?"
author: "Quynh Vo"
date: "2025-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is one of the projects from the Regression Analysis course in my third year undergrad. The analysis is conducted on the data set I was given in the course, thus I have no idea where the data comes from. I uploaded the data file (`spotify.csv`) in the same folder containing this Rmarkdown file.

Let's start with having a look at our data set.

```{r import}
#import data
df <- read.csv("spotify.csv")
head(df)
```

```{r check general info}
#look at the structure of df
str(df)
```
```{r check cleanliness}
#count missing values in each column
colSums(is.na(df))
```
Some preliminary steps shows that the given data set is clean and ready to use. There are 114000 observations representing songs on Spotify over 8 variables representing the recorded features of each song. All the variables are of numeric type, thus we can go ahead with looking at the summary statistics as well as some plots to grasp the general idea about the distribution of each song feature, which is part of the phase in the Data Lifecycle often referred as **Explanatory Data Analysis** (EDA).

```{r sum stats}
#summary statistics
summary(df)
```

```{r plot, echo=FALSE}
for (i in 1:length(colnames(df))) {
  #print(colnames(df)[i])
  plotName = cat("Distribution of ", colnames(df)[i])
  hist(data.matrix(df[colnames(df)[i]]),
       main = plotName,
       xlab = colnames(df)[i])
}
```

In general, this dataset looks fairly good and normalized, with features like `danceablity`, `energy`, `acousticness`, `instrumentalness`, `liveness` range between 0 and 1. `Popularity` seems to have been adjusted to the 0-100 scale. However, a potential issue stands out for `duration_ms` for some extreme outliers, as also pointed out in the histogram for `duration_ms`, (songs lasting over 5 million ms ≈ 87 minutes, which is unusual for typical songs).

```{r 1st model}
# Fit a linear model
model1 = lm(popularity ~ ., data=df)

# result 1
summary(model1)
```
First, the p-value obtained from the F-test is very small (< 2.2e-16), which shows that at least one of these predictors explain popularity, or overall the model does explain popularity. However, R-squared = 0.01 is very small, even after accounted for a large number of predictors, the adjusted R-squared = 0.0113 (~ 1.13%) shows the model explains little about the relationship of interest. This suggests a song popularity may be not be captured by these characteristics alone.

Second, looking closer to the p-values for each t-test, the intercept, danceability, energy, acousticness, instrumentalness and tempo statistically appear have a significant influence on a song's popularity. Specifically, energy, acousticness and instrumentalness have negative impact whereas the intercept, danceability and tempo have positive impact on popularity.

We got some good information so far, but still not very useful. We can try tweaking a little bit to see what else can we exploit from this model. We may guess that there may be some relationship between our predictors (for example, a song that is more acoustic tends to be less energetic, or danceable). Apparently, if it is really the case, we are violating the assumptions of a linear regression model (the predicting variables must be independent). Thus, let's have a check.

```{r 2nd model}
### Let's fit a second model without `acousticness`
model2 = lm(popularity ~ .-acousticness, data = df)

# result 2
summary(model2)

# change in coefficients
print("Coefficient shift: ")
model2$coefficients - model1$coefficients[-5] #omit `acousticness` in model 1
# coefficient shift: energy (-5.6 to -1.96 -> large increase), danceability (2.2 to 2.7)
```
First, we notice that djusted R-squared drops, this makes sense as we remove `acousticness` and it is a significant predictor. And then, `duration_ms` and `liveness`, which were insignificant before now become weakly significant. This may be an evidence for another guess about the relationship between `liveness` and `acousticness` since acoustic tracks are often recorded live.

Also, a closer look at the coefficients shows that removing `acousticness`, which has a negative effect on popularity, coefficient of `energy` increases. This shows `acousticness` and `energy` are correlated, verifying our guess about acoustic songs tend to be less energetic. Or by looking at the growth in `danceability` coefficient, we may guess `acousticness` was “masking” part of `danceability`’s effect since acoustic songs are often less danceable.

A more direct way to check for correlation between predictors is to use the correlation matrix.

```{r correlation matrix}
# correlation matrix
cor(subset(df, select = -popularity))
```

`energy` is shown to negatively correlate with `acousticness`.